\chapter{Charges in Electromagnetic Fields}\label{c3}
\begin{enumerate}
\item Unless they are in touch with each other, particles interact with each
other via fields.

\item Disturbances in fields propagate information. It is a postulate of the 
special theory of relativity that the changes in the state of a system are
progagated at a finite speed not exceeding $c$, the speed of light in vacuum.

\item This restriction rejects the hypothesis of an ideal rigid body in 
relativity. For suppose a force is applied to one end of a rigid body. Then in 
order that the inter-particle distance is unchanged, the force has to be 
propagated instantaneously to the other end. Therefore, in elementary analysis 
of the theory of relativity all particles are assumed to be points.

\item The principle of action for a free particle is given by equation 
\eqref{c2e4} and \eqref{c2e7},
\begin{equation}\label{c3e1}
S = -\int_{A}^{B}mcds,
\end{equation}
where $A$ and $B$ are world points.
For a particle in an electromagnetic field, the right hand side of this equation
should be augmented with another terms that represents the interaction of the
particle with the field. It should comprise of quantities describing the 
particle as well as the field.

The particle's interaction with the field is described by a single quantity, its
charge $e$ and the field by its 4-potential $A_\mu$. The complete action is
\begin{equation}\label{c3e2}
S = -\int_A^B \left(mcds + \frac{e}{c}A_\mu dx^\mu\right).
\end{equation}

The 4-potential can be written as 
\begin{equation}\label{c3e3}
A_\mu = (\varphi, \vec{A})
\end{equation}
so that we can write \eqref{c3e2} as
\begin{equation}\label{c3e4}
S = \int_A^B\left(-mcds - e\varphi dt + 
\frac{e}{c}\vec{A}\cdot d\vec{x}\right).
\end{equation}
Since $ds = c\sqrt{1 - \beta^2}dt$ and $d\vec{x} = \vec{v}dt$,
\begin{equation}\label{c3e5}
S = \int_{t_1}^{t_2}\left(-mc^2\sqrt{1 - \beta^2} - e\varphi + 
\frac{e}{c}\vec{A}\cdot\vec{v}\right)dt.
\end{equation}
Thus, the Lagrangian for a particle in an electromagnetic field is
\begin{equation}\label{c3e6}
L = -mc^2\sqrt{1 - \beta^2} - e\varphi + \frac{e}{c}\vec{A}\cdot\vec{v}
= -mc^2\sqrt{1 - \beta^2} - e\varphi + \frac{e}{c}A^i v_i.
\end{equation}
From the Lagrangian, we get the generalised momentum
\begin{equation}\label{c3e7}
p^i = \pdt{L}{v^i} = m\gamma v^i + \frac{e}{c}A^i.
\end{equation}
From equation \eqref{c2e8}, the first term is just the generalised momentum for
the Lagrangian of \eqref{c2e7}. The Hamiltonian is given be the Legendre
transform
\begin{eqnarray*}
\mathcal{H} &=& p^iv_i - L \\
  &=& m\gamma v^iv_i + \frac{e}{c}A^iv_i + mc^2\sqrt{1 - \beta^2} + e\varphi - 
  \frac{e}{c}A^iv_i \\
  &=& \frac{mv^2}{\sqrt{1 - \beta^2}} + mc^2\sqrt{1 - \beta^2} + e\varphi \\
  &=& \frac{mc^2}{\sqrt{1 - \beta^2}} + e\varphi
\end{eqnarray*}
or that
\begin{equation}\label{c3e8}
\mathcal{H} = mc^2\gamma + e\varphi.
\end{equation}
Stricly speaking \eqref{c3e8} is not the correct form of the Hamiltonian because
it is expressed in terms of the generalised velocity. To remedy this defect, we 
write
\eqref{c3e8} as
\[
\frac{(\mathcal{H} - e\varphi)^2}{c^2} = m^2c^2 \gamma^2.
\]
Now,
\begin{equation}\label{c3e9}
\gamma^2 = \frac{1}{1 - \beta^2} = 1 + \frac{\beta^2}{1 - \beta^2} = 
1 + \beta^2\gamma^2
\end{equation}
so that
\[
\frac{(\mathcal{H} - e\varphi)^2}{c^2} = m^2c^2(1 + \beta^2\gamma^2) 
\]
or that
\[
\frac{(\mathcal{H} - e\varphi)^2}{c^2} = m^2c^2 + m^2\gamma^2 v^2,
\]
From \eqref{c3e7} the last term on the rhs can be written as
\begin{equation}\label{c3e10}
\frac{(\mathcal{H} - e\varphi)^2}{c^2} = m^2c^2 + 
\left(p^i - \frac{e}{c}A^i\right)
\left(p_i - \frac{e}{c}A_i\right)
\end{equation}
Thus,
\begin{equation}\label{c3e11}
\mathcal{H} = e\varphi + \sqrt{m^2c^4 + c^2 
\left(p^i - \frac{e}{c}A^i\right)\left(p_i - \frac{e}{c}A_i\right)}
\end{equation}
is the correct form of the Hamiltonian. For speeds small as compared to the
speed of light,
\[
\sqrt{1 - \beta^2} = 1 - \frac{1}{2}\beta^2
\]
so that the Lagrangian of \eqref{c3e6} can be written as
\begin{equation}\label{c3e12}
L = \frac{1}{2}mv^2 - e\varphi + \frac{e}{c}\vec{A}\cdot\vec{v},
\end{equation}
where we ignored the constant terms $-mc^2$. At non-relativistic speeds, 
$\gamma$ can be well-approximated by $1$ so that the generalised momentum is
\begin{equation}\label{c3e13}
\vec{p} = m\vec{v} + \frac{e}{c}\vec{A}
\end{equation}
and the Hamiltonian is
\begin{equation}\label{c3e14}
\mathcal{H} = \frac{1}{2m}\left(\vec{p} - \frac{e}{c}\vec{A}\right)^2 + 
e\varphi.
\end{equation}

From \eqref{c2e27},
\[
p^\nu = \eta^{\mu\nu}p_\mu = -\eta^{\mu\nu}\pdt{S}{x^\mu}
\]
so that
\[
p^0 = -\frac{1}{c}\pdt{S}{t},\;\vec{p} = \grad S
\]
and hence the Hamilton-Jacobi equation is, from \eqref{c3e10},
\[
\frac{1}{c^2}\left(\pdt{S}{t} + e\varphi\right)^2 = m^2c^2 + 
\left(\grad S - \frac{e}{c}\vec{A}\right)^2
\]
where we replaced $\mathcal{H}$ in \eqref{c3e10} with $-\partial_t S$ 
because it is the energy of the particle and $p^0 = \mathcal{E}/c$. We can 
reexpress the relativistic Hamilton-Jacobi equation as 
\begin{equation}\label{c3e15}
\left(\grad S - \frac{e}{c}\vec{A}\right)^2 - 
\frac{1}{c^2}\left(\pdt{S}{t} + e\varphi\right)^2 + m^2c^2 = 0.
\end{equation}

\item The Lagrangian of \eqref{c3e12} allows us to derive the equation of 
motion.
\begin{eqnarray*}
\grad L &=& -e\grad\varphi + \frac{e}{c}(\vec{A}\cdot\grad\vec{v} + 
 \vec{v}\cdot\grad\vec{A} + \vec{A}\times\curl\vec{v} + 
 \vec{v}\times\curl\vec{A})\\ 
 &=& -e\grad\varphi + \frac{e}{c}(\vec{v}\cdot\grad\vec{A} + 
 \vec{v}\times\curl\vec{A})\\ 
\grad_{\vec{v}}L &=& m\vec{v} + \frac{e}{c}\vec{A}.
\end{eqnarray*}
The equation for $\grad L$ simplifies because $\vec{v}$ is independent of 
$\vec{x}$.  Therefore, $\grad\vec{v}$ and $\curl\vec{v}$ are both zero.
\begin{equation}\label{c3e16}
\frac{d}{dt}\left(m\vec{v} + \frac{e}{c}\vec{A}\right) = 
-e\grad\varphi + \frac{e}{c}(\vec{v}\cdot\grad\vec{A} + 
\vec{v}\times\curl\vec{A})
\end{equation}
is the equation of motion. Now.
\begin{equation}\label{c3e17}
\td{\vec{A}}{t} = \pdt{\vec{A}}{t} + \vec{v}\cdot\vec{A}
\end{equation}
so that \eqref{c3e16} is simplified to
\begin{equation}\label{c3e18}
\td{\vec{p}}{t} = e\left(-\grad\varphi - \frac{1}{c}\pdt{\vec{A}}{t}\right)
+ \frac{e}{c}\vec{v}\times\curl\vec{A}
\end{equation}
We define
\begin{eqnarray}
\vec{E} &=& -\grad\varphi - \frac{1}{c}\pdt{\vec{A}}{t} \label{c3e19} \\
\vec{H} &=& \curl\vec{A} \label{c3e20}
\end{eqnarray}
and call $\vec{E}$ the electric field and $\vec{H}$ the magnetic field. The
quantity
\begin{equation}\label{c3e21}
\vec{F} = e\left(\vec{E} + \frac{\vec{v}}{c}\times\vec{H}\right)
\end{equation}
is called the Lorentz force. This equation depends on the form \eqref{c3e12} of
the Lagrangian and \eqref{c3e13} of the momentum both of which are valid only in
the non-relativistic limit. As a result, \eqref{c3e18} to \eqref{c3e21} are 
applicable only in the non-relativistic limit. We derive the equation of motion
that is valid at all speeds later in equation \eqref{c3e74}. Interestingly, the
relativistic equation takes the same form as the non-relativistic one.

\item Recall that 
\begin{equation}\label{c3e22}
p^\mu = mcu^\mu = (mc\gamma, mv^i\gamma) = \left(\frac{\mathcal{E}}{c}, 
\vec{p}\right)
\end{equation}
so that the energy of the particle is
\begin{equation}\label{c3e23}
\mathcal{E} = mc^2\gamma
\end{equation}
so that
\begin{equation}\label{c3e24}
\td{\mathcal{E}}{t} = mc^2\td{\gamma}{t} = 
-\frac{mc^2}{2}\gamma^3\left(\frac{-2\vec{v}}{c^2}\cdot\td{\vec{v}}{t}\right)
= m\gamma^3 \vec{v}\cdot\td{\vec{v}}{t}
\end{equation}
and
\[
\td{\vec{p}}{t} = 
m\gamma\td{\vec{v}}{t} + 
m\gamma^3\frac{\vec{v}}{c^2}\vec{v}\cdot\td{\vec{v}}{t} = 
m\gamma^3\left(m\gamma^{-2}\td{\vec{v}}{t} + 
\frac{\vec{v}}{c^2}\vec{v}\cdot\td{\vec{v}}{t}\right)
\]
or, since
\[
\gamma = \frac{1}{\sqrt{1 - v^2/c^2}},
\]
we have,
\[
\td{\vec{p}}{t} = m\gamma^3\left(
\left(1 - \frac{v^2}{c^2}\right)\td{\vec{v}}{t} + 
\frac{\vec{v}}{c^2}\vec{v}\cdot\td{\vec{v}}{t}\right)
\]
Therefore,
\begin{equation}\label{c3e25}
\vec{v}\cdot\td{\vec{p}}{t} = 
m\gamma^3\left(\left(1 - \frac{v^2}{c^2}\right)\vec{v}\cdot\td{\vec{v}}{t} + 
\frac{v^2}{c^2}\vec{v}\cdot\td{\vec{v}}{t}\right) = 
m\gamma^3\vec{v}\cdot\td{\vec{v}}{t}
\end{equation}
From equations \eqref{c3e24} and \eqref{c3e25},
\begin{equation}\label{c3e26}
\td{\mathcal{E}}{t} = \vec{v}\cdot\td{\vec{p}}{t}.
\end{equation}
If we apply it to the equation of motion \eqref{c3e18} and use the definitions 
of the fields, we get
\begin{equation}\label{c3e27}
\td{\mathcal{E}}{t} = e\vec{v}\cdot\vec{E}.
\end{equation}

Here we have used the fact that
\begin{equation}\label{c3e28}
\td{v^2}{t} = \td{(\vec{v}\cdot\vec{v})}{t^2} = 2\vec{v}\cdot\td{\vec{v}}{t}.
\end{equation}

Equation \eqref{c3e27} suggests that the energy of a system of charged 
particles changes only because of the electric field. The magnetic field does 
not change the energy.

\item The equation of motion
\begin{equation}\label{c3e29}
\td{\vec{p}}{t} = e\left(\vec{E} + \frac{\vec{v}}{c}\times\vec{H}\right)
\end{equation}
is invariant under time reversal if we also reverse the direction of the 
magnetic field.

\item A 4-potential $A_\mu$ gives a unique combination of electric and magnetic
fields. But the relationship is many-to-one. This is because, the potential
\begin{equation}\label{c3e30}
A_\mu^\op = A_\mu - \pdt{f}{x^\mu}
\end{equation}
results in
\begin{eqnarray}
\varphi^\op &=& \varphi - \frac{1}{c}\pdt{f}{t} \label{c3e31} \\
\vec{A}^\op &=& \vec{A} + \grad f \label{c3e32}
\end{eqnarray}
and the fields
\begin{eqnarray}
\vec{E}^\op &=& 
	-\grad{\varphi^\op} - \frac{1}{c}\pdt{\vec{A}^\op}{t} \nonumber \\
 &=& -\grad\varphi + \frac{1}{c}\grad\pdt{f}{t} - 
      \frac{1}{c}\pdt{\vec{A}}{t} - \frac{1}{c}\grad{\pdt{f}{t}} \nonumber \\
 &=& -\grad\varphi - \frac{1}{c}\pdt{\vec{A}}{t} \nonumber \\
 &=& \vec{E} \label{c3e33} \\
\vec{H}^\op &=& \curl\vec{A}^\op \nonumber \\
 &=& \curl\vec{A} \nonumber \\
 &=& \vec{H} \label{c3e34}
\end{eqnarray}
Equation \eqref{c3e30} is called the gauge transformation and an invariance of
functions under it is called gauge invariance. All physically meaningful 
quantities must be gauge invariant.

\item If $A_\mu$ is independent of time then so are the fields. $\varphi$
influences only the electric field Further, since the Lagrangian is 
independent of time, the energy of the system is a constant and it coincides 
with the Hamiltonian. From
\eqref{c3e8}, it is
\begin{equation}\label{c3e35}
\mathcal{E} (= \mathcal{H}) = mc^2\gamma + e\varphi.
\end{equation} 
The effect of the fields is just to add a term $e\varphi$ to the energy. This 
term is called the \emph{potential energy}. It is independent of $\vec{A}$ 
which means that the magnetic field has no influence on the particle's energy.

If the fields are constant then one can easily verify that
\begin{eqnarray}
\varphi &=& -\vec{E}\cdot\vec{x} \label{c3e36} \\
\vec{A} &=& \frac{1}{2}\vec{H} \times \vec{x} \label{c3e37}
\end{eqnarray}

\item We will now consider the motion of a charge particle in a constant 
electric field. From the equation of motion \eqref{c3e29},
\begin{equation}\label{c3e38}
\td{\vec{p}}{t} = e\vec{E}.
\end{equation}
If $\vec{E} = E\uv{x}$ then we have
\begin{eqnarray}
p_x(t) &=& p_x(0) + eEt \label{c3e39} \\
p_y(t) &=& p_y(0) \label{c3e40}.
\end{eqnarray}
Let us choose the zero of time at an instant when $p_x(0)=0$ and $p_y(0)=p_0$,
the initial momentum of the particle. Since $\vec{p} = m\gamma\vec{v}$ and 
$\mathcal{E} = m\gamma c^2$, we have 
\begin{equation}\label{c3e41}
\vec{v} = \frac{\vec{p}c^2}{\mathcal{E}}
\end{equation}
so that
\begin{eqnarray}
\dot{x}(t) &=& \frac{ec^2Et}{\mathcal{E}} \label{c3e42} \\
\dot{y}(t) &=& \frac{c^2p_0}{\mathcal{E}} \label{c3e43}
\end{eqnarray}
A particle in an electric field gains energy. Therefore, $\mathcal{E}$ is not a
constant and hence \eqref{c3e42} cannot be integrated as is. However, we know 
that $\mathcal{E}^2 = m^2c^4 + p^2c^2 = m^2c^4 + p_0^2c^2 + e^2E^2t^2c^2$ so 
that equation \eqref{c3e42} becomes
\begin{equation}\label{c3e44}
\dot{x}(t) = \frac{ec^2Et}{\sqrt{m^2c^4 + p_0^2c^2 + c^2e^2E^2t^2}}.
\end{equation}
and hence
\[
x(t) = \int \frac{ec^2Et dt}{\sqrt{m^2c^4 + p_0^2c^2 + c^2e^2E^2t^2}} + x(0).
\]
If $ceEt = u, dt = du/(ceE)$ and
\[
x(t) = \frac{1}{eE}\int\frac{u du}{\sqrt{\mathcal{E}_0^2 + u^2}} + x(0)
\]
If $u = \mathcal{E}_0\sinh v$, $du = \mathcal{E}_0\cosh  vdv$ and
\begin{eqnarray*}
x(t) &=& 
 \frac{1}{eE}\int\frac{\mathcal{E}_0^2\sinh v\cosh v dv}
 	{\mathcal{E}_0\cosh v} + x(0) \\
 &=& \frac{\mathcal{E}_0}{eE}\int\sinh v dv + x(0) \\
 &=& \frac{\mathcal{E}_0}{eE}\cosh v + x(0) \\
 &=& \frac{1}{eE}\sqrt{\mathcal{E}_0^2 + u^2} + x(0)
\end{eqnarray*}
or
\begin{equation}\label{c3e45}
x(t) = \frac{1}{eE}\sqrt{\mathcal{E}_0^2 + e^2c^2E^2t^2} + x(0)
\end{equation}
Analogous to \eqref{c3e44}
\[
\dot{y}(t) = \frac{p_0c^2}{\sqrt{\mathcal{E}_0^2 + e^2E^2c^2t^2}}
\]
so that
\[
y(t) = \int \frac{p_0c^2}{\sqrt{\mathcal{E}_0^2 + e^2E^2c^2t^2}} dt + y(0).
\]
Put $ecEt = \mathcal{E}_0\sinh u$ so that $ecEdt = \mathcal{E}_0\cosh u du$
and
\[
y(t) = \frac{p_0c}{eE}\int \frac{\mathcal{E}_0\cosh u du}{\mathcal{E}_0\cosh u}
 + y(0) = \frac{p_0c}{eE} u + y(0) 
\]
that is
\begin{equation}\label{c3e46}
y(t) = \frac{p_0c}{eE} \sinh^{-1}\left(\frac{ecEt}{\mathcal{E}_0}\right) + y(0).
\end{equation}
We choose the origin such that $x(0) = 0, y(0) = 0$ so that from \eqref{c3e46}
we have
\[
ecEt = \mathcal{E}_0\sinh\left(\frac{eEy}{p_0c}\right)
\]
Substituting this in \eqref{c3e45} we get the equation of the orbit
\begin{equation}\label{c3e47}
x(y) = \frac{\mathcal{E}_0}{eE}\cosh\left(\frac{eEy}{p_0c}\right).
\end{equation}
It describes a catenary symmetric about the $x$-axis. When $eEy \ll p_0c$, we
can approximate \eqref{c3e47} as
\begin{equation}\label{c3e48}
x(y) = \frac{\mathcal{E}_0}{eE}
	\left(1 + \frac{1}{2}\frac{e^2E^2y^2}{p_0^2c^2}\right),
\end{equation}
which is an equation of a parabola.

\item When a charged particle is subjected to a constant magnetic field, the
equation of motion is
\begin{equation}\label{c3e49}
\td{\vec{p}}{t} = e\frac{\vec{v}}{c} \times \vec{H}.
\end{equation}
Once again, we use the fact that $\mathcal{E} = m\gamma c^2$ and $\vec{p} = 
m\gamma\vec{v}$ to get
\[
\vec{p} = \frac{\mathcal{E}}{c^2}\vec{v}
\]
so that \eqref{c2e48} becomes
\[
\td{\vec{v}}{t} = \frac{ec}{\mathcal{E}}\vec{v} \times \vec{H}.
\]
In a magnetic field, the energy of a particle remains unchanged. Therefore, we 
could pull $\mathcal{E}$ out of the integral. Let us align the $z$ axis along
the magnetic field so that $\vec{H} = H\uv{z}$ and $\vec{v} \times \vec{H} = 
\uv{x}(v_yH - 0) + \uv{y}(-v_xH) + \uv{z}(0)$. Therefore,
\begin{eqnarray*}
\dot{v}_x(t) &=& \frac{ecH}{\mathcal{E}}v_y \\
\dot{v}_y(t) &=& -\frac{ecH}{\mathcal{E}}v_x \\
\dot{v}_z(t) &=& 0
\end{eqnarray*}
From these equations, we get
\begin{eqnarray*}
\ddot{v}_x(t) &=& -\omega^2 v_x \\
\ddot{v}_y(t) &=& -\omega^2 v_y \\
\ddot{v}_z(t) &=& 0
\end{eqnarray*}
where
\begin{equation}\label{c3e50}
\omega = \frac{ecH}{\mathcal{E}},
\end{equation}
is called the cyclotron frequency. At low speeds, $\gamma \approx 1$, 
$\mathcal{E} = mc^2$ and the cyclotron frequency is $(eH)/(mc)$. Note that 
$\omega$ is a constant. We can write solutions to the differential equations 
of velocity components as
\begin{eqnarray*}
v_x(t) &=& A_1\omega\cos\omega t + A_2\omega\sin\omega t \\
v_y(t) &=& A_3\omega\cos\omega t + A_4\omega\sin\omega t \\
v_z(t) &=& v_z(0)
\end{eqnarray*}
here $A_1, A_2, A_3, A_4$ are the constants of integration. If the initial
conditions are $v_x(0) = v_{\perp}, v_y(0) = 0$ then $A_1 = v_{\perp}$ and 
$A_3 = 0$ so that
\begin{eqnarray*}
v_x(t) &=& v_\perp\cos\omega t + A_2\omega\sin\omega t \\
v_y(t) &=& A_4\omega\sin\omega t \\
v_z(t) &=& v_z(0)
\end{eqnarray*}
from which we get
\begin{eqnarray*}
x(t) &=& \frac{v_\perp}{\omega}\sin\omega t + A_2\cos\omega t \\
y(t) &=& -A_4\cos\omega t \\
z(t) &=& z(0) + v_z(0)t
\end{eqnarray*}
If $x(0) = 0, y(0) = r$ then $A_2 = 0$ and $A_4 = -r$ so that
\begin{eqnarray}
x(t) &=& \frac{v_\perp}{\omega}\sin\omega t \label{c3e51} \\
y(t) &=& r\cos\omega t \label{c3e52} \\
z(t) &=& z(0) + v_z(0)(t) \label{c3e53}
\end{eqnarray}
The constants $v_\perp$ and $r$ are not unrelated to each other. Using them in
the equations of velocity components we get
\begin{eqnarray*}
v_x(t) &=& v_\perp\cos\omega t \\
v_y(t) &=& -r\omega\sin\omega t
\end{eqnarray*}
so that $\dot{v}_x(t) = v_\perp\omega\sin\omega t$. From the equation of motion
the right hand side is $\omega v_y = -r\omega^2\sin\omega t$ so that we get
\begin{equation}\label{c3e54}
v_\perp = -r\omega.
\end{equation}
The equations of the trajectory therefore simplify to
\begin{eqnarray}
x(t) &=& -r\sin\omega t \label{c3e55} \\
y(t) &=& r\cos\omega t \label{c3e56} \\
z(t) &=& z(0) + v_z(0)(t) \label{c3e57}
\end{eqnarray}
This is an equation of an helix.

\item Now consider a charged particle in the presence of a constant magnetic 
field $\vec{H} = H\uv{z}$ and a constant electric field $\vec{E} = E_y\uv{y} + 
E_z\uv{z}$. We restrict ourselves to non-relativistic regime so that $\vec{p} 
= m\vec{v}$ and the equation of motion \eqref{c3e29} becomes
\begin{eqnarray}
m\dot{v}_x &=& \frac{e}{c}v_yH \label{c3e58} \\
m\dot{v}_y &=& eE_y - \frac{e}{c}v_xH \label{c3e59} \\
m\dot{v}_z &=& eE_z \label{c3e60}
\end{eqnarray}
The last equation immediately gives
\[
v_z = \frac{e}{m}E_zt + v_z(0)
\]
and
\begin{equation}\label{c3e61}
z(t) = \frac{e}{2m}E_zt^2 + v_z(0)t + z(0).
\end{equation}
In order to solve the coupled equations \eqref{c3e57} and \eqref{c3e58}, we can 
either differentiate them once more and decouple them or use a different trick.
We demonstrated the former in the previous point. Here, we will use the latter.
Let $\tilde{v} = v_x + iv_y$ so that
\[
m\td{\tilde{v}}{t} = \frac{e}{c}v_yH + ieE_y - \frac{e}{c}{iv_xH}
= ieE_y -i\frac{e}{c}(H(v_x + iv_y)) = ieE_y - i\frac{eH}{c}\tilde{v}
\]
or
\begin{equation}\label{c3e62}
\td{\tilde{v}}{t} = -i\frac{eH}{mc}\tilde{v} + i\frac{e}{m}E_y,
\end{equation}
It can be readily integrated to
\[
\ln\left(\tilde{v} - \frac{eE_y}{m\omega}\right) = -i\omega t + \ln\tilde{v}(0)
\]
or
\begin{equation}\label{c3e63}
\tilde{v} = \tilde{v}(0)e^{-i\omega t} + \frac{eE_y}{m\omega},
\end{equation}
where we approximated \eqref{c3e50} for non-relativistic speeds. Equating the
real and imaginary parts,
\begin{eqnarray}
v_x(t) &=& v_x(0)\cos\omega t  + \frac{e}{m\omega}E_y\label{c3e64} \\
v_y(t) &=& -v_y(0)\sin\omega t \label{c3e65}
\end{eqnarray}
From these equations, we readily get
\begin{eqnarray}
\langle v_x(t) \rangle &=& \frac{e}{m\omega}E_y \label{c3e66} \\
\langle v_y(t) \rangle &=& 0 \label{c3e67}
\end{eqnarray}
Thus, the motion remains bounded on the $y$ axis but not so on the other axes. 
Equations \eqref{c3e65} and \eqref{c3e66} can be readily integrated to
\begin{eqnarray}
x(t) &=& 
	\frac{v_x(0)}{\omega}\sin\omega t + \frac{e}{m\omega}E_yt + x(0) 
	\label{c3e68} \\
y(t) &=& \frac{v_y(0)}{\omega}\cos\omega t + y(0) \label{c3e69}
\end{eqnarray}
Equations \eqref{c3e61}, \eqref{c3e68} and \eqref{c3e69} are the equations of 
the trajectory.

\item We derived the equations of motion in point (4) from the Lagrangian which
was obtained after writing the principle of least action as an integral over 
time. We can, instead, apply the variational method to equation \eqref{c3e2}. 
Thus,
\begin{eqnarray*}
\delta S &=& -\delta\int_A^B\left(mcds + \frac{e}{c}A_\mu dx^\mu\right) \\
 &=& -\int_A^B\left(mc\delta ds + \frac{e}{c}\delta(A_\mu dx^\mu)\right) \\
 &=& -\int_A^B\left(mc\delta\sqrt{dx_\mu dx^\mu} + 
 	\frac{e}{c}(\delta(A_\mu) dx^\mu + A_\mu d\delta x^\mu)\right) \\
 &=& -\int_A^B\left(mc\frac{1}{2}
 	\frac{dx_\mu d\delta x^\mu + d(\delta x_\mu)dx^\mu}{ds}
     + \frac{e}{c}(\delta(A_\mu) dx^\mu + A_\mu d\delta x^\mu)\right)
\end{eqnarray*}
Now $d(\delta x_\mu)dx^\mu = d(\delta x^\mu)dx_\mu = dx_\mu d(\delta x^\mu)$ so 
that
\begin{equation}\label{c3e70}
\delta S = -\int_A^B\left(mc\frac{dx_\mu d(\delta x^\mu)}{ds}
     + \frac{e}{c}(\delta(A_\mu) dx^\mu + A_\mu d\delta x^\mu)\right)
\end{equation}    
Using \eqref{c1e78},
\begin{equation}\label{c3e71}
\delta S = -\int_A^B\left((mcu_\mu + \frac{e}{c}A_\mu)d\delta x^\mu + 
\frac{e}{c}\delta(A_\mu) dx^\mu\right).
\end{equation}
We can integrate first term by parts,
\[
\delta S = -\left(mcu_\mu + \frac{e}{c}A_\mu\right)\delta x^\mu\Big|_A^B + 
 \int_A^B \left(\delta x^\mu d\left(mcu_\mu + \frac{e}{c}A_\mu\right) - 
 \frac{e}{c}\delta(A_\mu) dx^\mu\right)
\]
Since the variations $\delta x^\mu$ vanish at the end points,
\[
\delta S = \int_A^B\left(\delta x^\mu\left(mcdu_\mu + \frac{e}{c}dA_\mu\right) 
- \frac{e}{c}\delta(A_\mu) dx^\mu\right)
\]
We now write
\[
dA_\mu = \pdt{A_\mu}{x^\nu}dx^\nu;\; \delta A_\mu = 
\pdt{A_\mu}{x^\nu}\delta x^\nu 
\]
so that
\begin{eqnarray*}
\delta S &=& 
	\int_A^B\left(mcdu_\mu\delta x^\mu + 
	\frac{e}{c}\pdt{A_\mu}{x^\nu}dx^\nu\delta x^\mu
 - \frac{e}{c}\pdt{A_\mu}{x^\nu}\delta x^\nu dx^\mu\right) \\
 &=& \int_A^B\left(mcdu_\mu\delta x^\mu + 
 	\frac{e}{c}\pdt{A_\mu}{x^\nu}dx^\nu\delta x^\mu
 - \frac{e}{c}\pdt{A_\nu}{x^\mu}\delta x^\mu dx^\nu\right) \\
 &=& \int_A^B\left(mcdu_\mu - \frac{e}{c}\left(\pdt{A_\nu}{x^\mu}
 - \pdt{A_\mu}{x^\nu}\right)dx^\nu\right)\delta x^\mu \\
 &=& \int_A^B\left(mc\td{u_\mu}{s}ds - \frac{e}{c}\left(\pdt{A_\nu}{x^\mu}
 - \pdt{A_\mu}{x^\nu}\right)u^\nu ds\right)\delta x^\mu \\
 &=& \int_A^B\left(mc\td{u_\mu}{s} - \frac{e}{c}\left(\pdt{A_\nu}{x^\mu}
 - \pdt{A_\mu}{x^\nu}\right)u^\nu \right)ds\delta x^\mu
\end{eqnarray*}
$\delta S = 0$ implies that
\[
\int_A^B\left(mc\td{u_\mu}{s} - \frac{e}{c}\left(\pdt{A_\nu}{x^\mu}
 - \pdt{A_\mu}{x^\nu}\right)u^\nu \right)ds\delta x^\mu = 0.
\]
Since this is true for all $\delta x^\mu = 0$,
\[
\int_A^B\left(mc\td{u_\mu}{s} - \frac{e}{c}\left(\pdt{A_\nu}{x^\mu}
 - \pdt{A_\mu}{x^\nu}\right)u^\nu \right)ds = 0.
\]
This integral is always zero, we must have
\begin{equation}\label{c3e72}
mc\td{u_\mu}{s} = 
\frac{e}{c}\left(\pdt{A_\nu}{x^\mu} - \pdt{A_\mu}{x^\nu}\right)u^\nu
\end{equation}
Define the electromagnetic field tensor
\begin{equation}\label{c3e73}
F_{\mu\nu} = \pdt{A_\nu}{x^\mu} - \pdt{A_\mu}{x^\nu}
\end{equation}
so that the equation of motion becomes
\[
mc\td{u_\mu}{s} = \frac{e}{c}F_{\mu\nu}u^\nu.
\]
equivalently,
\begin{equation}\label{c3e74}
mc\td{u^\mu}{s} = \frac{e}{c}F^{\mu\nu}u_\nu.
\end{equation}
Recall that $dx^\mu = (cdt, dx^1, dx^2, dx^3)$ so that $dx_\mu = (cdt, -dx^1, 
-dx^2, -dx^3)$. Further, $A_\mu = (\varphi, -A_1, -A_2, -A_3)$ so that the 
components of $F_{\mu\nu}$ are
\[
\begin{bmatrix}
0 & -c^{-1}\partial_t A_1 - \partial_{x^1}\varphi & 
-c^{-1}\partial_t A_2 - \partial_{x^2}\varphi & 
`-c^{-1}\partial_t A_3 - \partial_{x^3}\varphi \\
c^{-1}\partial_t A_1 - \partial_{x^1}\varphi & 0 & 
-\partial_{x^1}A_2+\partial_{x^2}A^1 & -\partial_{x^1}A_3+\partial_{x^3}A^3\\
c^{-1}\partial_t A_2 - \partial_{x^2}\varphi & 
\partial_{x^1}A_2-\partial_{x^2}A_1 & 0 & 
-\partial_{x^2}A_3+\partial_{x^3}A_2\\
c^{-1}\partial_t A_3 - \partial_{x^3}\varphi & 
\partial_{x^1}A_3-\partial_{x^3}A_1 & \partial_{x^3}A_2-\partial_{x^3}A_2 & 0
\end{bmatrix}
\]
From equations \eqref{c3e19} and \eqref{c3e20},
\begin{equation}\label{c3e75}
F_{\mu\nu} = 
\begin{bmatrix}
0 & E_1 & E_2 & E_3 \\
-E_1 & 0 & -H_3 & H_2\\
-E_2 & H_3 & 0 & -H_1\\
-E_3 & -H_2 & H_1 & 0
\end{bmatrix}
\end{equation}
In order to transform this into a contravariant tensor, we use the equation
\begin{equation}\label{c3e76}
F^{\mu\nu} = \eta^{\mu\sigma}F_{\sigma\rho}\eta^{\rho\nu}.
\end{equation}
The factors are written in a manner that mimics the matrix multiplication. The 
matrix representing the metric is $\text{diag}(1, -1, -1, -1)$ so that the 
result is
\begin{equation}\label{c3e77}
F^{\mu\nu} = 
\begin{bmatrix}
0 & -E_1 & -E_2 & -E_3 \\
E_1 & 0 & -H_3 & H_2\\
E_2 & H_3 & 0 & -H_1\\
E_3 & -H_2 & H_1 & 0
\end{bmatrix}
\end{equation}
Note that $E_i, H_i$ are not space components of a 4-vector. They should be 
treated as scalars. That is $E_i = E^i$ and $H_i = H^i$.


\item Let us write \eqref{c3e74} in components form. Since
\[
u^\mu = 
\left(\gamma, -\frac{\gamma}{c}v^1, -\frac{\gamma}{c}v^2, 
-\frac{\gamma}{c}v^3\right),
\]
we have
\[
c\td{u^\mu}{s} = \left(c\td{\gamma}{s}, -\td{(\gamma v^1)}{s}, 
-\td{(\gamma v^2)}{s}, -\td{(\gamma v^3)}{s}\right)
\]
and
\[
u_\mu = \left(\gamma, \frac{\gamma}{c}v_1, \frac{\gamma}{c}v_2, 
\frac{\gamma}{c}v_3\right),
\]
and hence
\begin{eqnarray*}
mc\td{\gamma}{s} &=& \frac{e}{c}(F^{00}\gamma - F^{01}\gamma \beta_1 - 
F^{02}\gamma \beta_2 -F^{03}\gamma \beta_3) \\
-m\frac{d}{ds}(\gamma v^1)&=&\frac{e}{c}(F^{10}\gamma + F^{11}\gamma \beta_1 +
F^{12}\gamma \beta_2 + F^{13}\gamma \beta_3) \\
-m\frac{d}{ds}(\gamma v^2)&=&\frac{e}{c}(F^{20}\gamma + F^{21}\gamma \beta_1 + 
F^{22}\gamma \beta_2 + F^{23}\gamma \beta_3) \\
-m\frac{d}{ds}(\gamma v^3)&=&\frac{e}{c}(F^{30}\gamma + F^{31}\gamma \beta_1 + 
F^{32}\gamma \beta_2 + F^{33}\gamma \beta_3) 
\end{eqnarray*}
Since $v^i = -v_i$ and $ds = cdt/\gamma$,
\begin{eqnarray*}
mc\td{\gamma}{t} &=& e(E_1 \beta_1 + E_2 \beta_2 + E_3 \beta_3) \\
m\frac{d}{dt}(\gamma v_1) &=& e(E_1 + H_3 \beta_2 - H_2 \beta_3) \\
m\frac{d}{dt}(\gamma v_2) &=& e(E_2 - H_3 \beta_1 + H_1 \beta_3) \\
m\frac{d}{dt}(\gamma v_3) &=& e(E_3 + H_2 \beta_1 - H_1 \beta_2).
\end{eqnarray*}
The last three equations can be written in vector form as
\[
\frac{d}{dt}(m\gamma\vec{v}) = \td{\vec{p}}{t} = 
e(\vec{E} + \vec{\beta} \times \vec{H})
\]
which is indeed the same as \eqref{c3e29}. The first equation can be written as
\[
mc^2\td{\gamma}{t} = \frac{d}{dt}(mc^2\gamma) = 
\td{\mathcal{E}}{t} = e\vec{E}\cdot\vec{v}
\]
This is the ``relativistic'' form of the work-energy theorem of \eqref{c3e27}.

\item From the equation of motion \eqref{c3e74} we also have
\[
mcu_\mu\td{u^\mu}{s} = \frac{e}{c}u_\mu F^{\mu\nu}u_\nu.
\]
The left hand side is zero because of \eqref{c1e86} while the right hand side 
vanishes because of anti-symmetry of the electromagnetic field tensor. This 
means that the four equations of motion in the previous point are not 
independent.

\item We go back to equation \eqref{c3e71}
\[
\delta S = -\int_A^B\left((mcu_\mu + \frac{e}{c}A_\mu)d\delta x^\mu + 
\frac{e}{c}\delta(A_\mu) dx^\mu\right).
\]
and its immediate consequence,
\begin{eqnarray*}
\delta S &=& -\left(mcu_\mu + \frac{e}{c}A_\mu\right)\delta x^\mu\Big|_A^B + \\
  & & \int_A^B \left(\delta x^\mu d\left(mcu_\mu + \frac{e}{c}A_\mu\right) - 
 \frac{e}{c}\delta(A_\mu) dx^\mu\right)
\end{eqnarray*}
The first term is a difference of a quantity at two fixed points while the 
second one is a variation over all paths between those points. Since the 
variations vanish at the fixed points the first term is zero while the second 
term, after some manipulations, gives the equations of motion.

If we consider only the actual paths starting from the world-point $A$ and let 
$H$ be any other point on it, then the second term will necessarily be zero 
while the first one will not be zero. The equation then gives,
\[
\frac{\delta S}{\delta x^\mu} = -\left(mcu_\mu + \frac{e}{c}A_\mu\right).
\]
In the limit $\delta x^\mu \rightarrow 0$ this equation becomes
\begin{equation}\label{c3e78}
-\pdt{S}{x^\mu} = mcu_\mu + \frac{e}{c}A_\mu.
\end{equation}
From equation \eqref{c2e26}, the left hand side is the generalised momentum
\begin{equation}\label{c3e79}
p_\mu = mcu_\mu + \frac{e}{c}A_\mu.
\end{equation}
This agrees with \eqref{c3e7}.

\item From equation \eqref{c1e46}, the Lorentz transformation of the 4-potential
gives
\begin{equation}\label{c3e80}
\varphi = \gamma(\bar{\varphi} + \beta\bar{A}_1);
\; A_1 = \gamma(\bar{A}_1 + \beta\bar{\varphi}); A_2 = \bar{A}_2;\; 
A_3 = \bar{A}_3.
\end{equation}
Using the solution to problem 2 in chapter \ref{c1}, we have $F^{01} = 
\bar{F}^{01}$ (from \eqref{c1e98}) so that 
\begin{equation*}
E_1 = \bar{E}_1.
\end{equation*}
From \eqref{c1e90}
\[
F^{02} = \frac{\bar{F}^{02} + \beta\bar{F}^{12}}{1 - \beta^2}
\]
so that
\begin{equation*}
E_2 = \frac{\bar{E}_2 + \beta\bar{H}_3}{1 - \beta^2}.
\end{equation*}
From \eqref{c1e91}
\[
F^{03} = \frac{\bar{F}^{03} + \beta\bar{F}^{13}}{1 - \beta^2}
\]
so that
\begin{equation*}
E_3 = \frac{\bar{E}_3 - \beta\bar{H}_2}{1 - \beta^2}.
\end{equation*}
From \eqref{c1e93}
\[
F^{12} = \frac{\bar{F}^{12} + \beta\bar{F}^{02}}{1 - \beta^2}
\]
so that
\begin{equation*}
-H_3 = \frac{-\bar{H}_3 - \beta\bar{E}_2}{1 - \beta^2} \Rightarrow 
H_3 = \frac{\bar{H}_3 + \beta\bar{E}_2}{1 - \beta^2}
\end{equation*}
From \eqref{c1e94}
\[
F^{13} = \frac{\bar{F}^{13} + \beta\bar{F}^{03}}{1 - \beta^2}
\]
so that
\begin{equation*}
H_2 = \frac{\bar{H}_2 - \beta\bar{E}_3}{1 - \beta^2}.
\end{equation*}
Finally, $F^{23} = \bar{F}^{23}$ implies
\begin{equation*}
H_1 = \bar{H}_1.
\end{equation*}
We collect these formulae together as
\begin{eqnarray}
E_1 &=& \bar{E}_1 \\ \label{c3e81}
E_2 &=& \frac{\bar{E}_2 + \beta\bar{H}_3}{1 - \beta^2} \\ \label{c3e82}
E_3 &=& \frac{\bar{E}_3 - \beta\bar{H}_2}{1 - \beta^2} \\ \label{c3e83}
H_1 &=& \bar{H}_1 \\ \label{c3e84}
H_2 &=& \frac{\bar{H}_2 - \beta\bar{E}_3}{1 - \beta^2} \\ \label{c3e85}
H_3 &=& \frac{\bar{H}_3 + \beta\bar{E}_2}{1 - \beta^2} \label{c3e86}
\end{eqnarray}

\item At low speeds, $1 - \beta^2 \approx 1$ and the transformation equations
become
\begin{eqnarray}
E_1 &=& \bar{E}_1 \\ \label{c3e87}
E_2 &=& \bar{E}_2 + \beta\bar{H}_3 \\ \label{c3e88}
E_3 &=& \bar{E}_3 - \beta\bar{H}_2 \\ \label{c3e89}
H_1 &=& \bar{H}_1 \\ \label{c3e90}
H_2 &=& \bar{H}_2 - \beta\bar{E}_3 \\ \label{c3e92}
H_3 &=& \bar{H}_3 + \beta\bar{E}_2 \label{c3e93}
\end{eqnarray}
Since $\beta = v/c$ or really $v_1/c$, we can write these equations in vector 
form as
\begin{eqnarray}
\vec{E} = \vec{E}^\op + \frac{1}{c}\vec{H}^\op \times \vec{v} \label{c3e94} \\
\vec{H} = \vec{H}^\op - \frac{1}{c}\vec{E}^\op \times \vec{v}, \label{c3e95}
\end{eqnarray}
where $\vec{E}^\op = (\bar{E}_1, \bar{E}_2, \bar{E}_3)$ and $\vec{H}^\op = 
(\bar{H}_1, \bar{H}_2, \bar{H}_3)$, These equations tell that even if 
$\vec{H}^\op = 0$ ($\vec{E}^\op = 0$), $\vec{H}$ ($\vec{E}$) is not. Further, 
it is clear that $\vec{E}$ is perpendicular to $\vec{H}$ in both cases.

\item Suppose $\vec{E}^\op$ and $\vec{H}^\op$ are perpendicular to each other.
If $\vec{v} = c\vec{H}^\op/E^\op$ then, from \eqref{c3e93} and \eqref{c3e94},
\begin{eqnarray}
\vec{E} &=& \vec{E}^\op \label{c3e96} \\
\vec{H} &=& 0 \label{c3e97}
\end{eqnarray}
Similarly, if we choose $\vec{v} = c\vec{E}^\op/H^\op$, we will get $\vec{E} = 
0$ and $\vec{H} = \vec{H}^\op$. Thus, if $\vec{E}^\op$ and $\vec{H}^\op$ are 
perpendicular to each other then one can find a frame of reference in which one
of the fields is zero.

\item For any tensor, $A^{\mu\nu}$, $A_{\mu\nu}A^{\mu\nu}$ and 
$\epsilon^{\mu\rho\nu\sigma}A^{\mu\rho}A^{\nu\sigma}$ are scalars, and therefore
invariants under Lorentz transformation. In the case of electromagnetic field
tensor, the first of these is
\[
F_{\mu\nu}F^{\mu\nu} = -E_1^2 - E_2^2 - E_3^2 + H_3^2 + H_2^2 + H_1^2.
\]
Thus, the quantity 
\begin{equation}\label{c3e98}
H^2 - E^2
\end{equation}
is an invariant. To find the next invariant, we
need the product
\begin{eqnarray*}
\epsilon^{\mu\rho\nu\sigma}F^{\mu\rho}F^{\nu\sigma} &=& 
	 F^{01}F^{23} - F^{01}F^{32} - F^{02}F^{13} + F^{02}F^{31} \\
 & & F^{03}F^{12} - F^{03}F^{21} - F^{10}F^{23} + F^{10}F^{32} \\
 & & F^{12}F^{03} - F^{12}F^{30} - F^{13}F^{02} + F^{13}F^{20} \\
 & & F^{20}F^{13} - F^{20}F^{31} - F^{21}F^{03} + F^{21}F^{30} \\
 & & F^{23}F^{01} - F^{23}F^{10} - F^{30}F^{12} + F^{30}F^{21} \\
 & & F^{31}F^{02} - F^{31}F^{20} - F^{32}F^{01} + F^{32}F^{10} \\
 &=& E_1H_1 + E_1H_1 + E_2H_2 + E_2H_2 \\
 & & E_3H_3 + E_3H_3 + E_1H_1 + E_1H_1 \\
 & & E_3H_3 + E_3H_3 + E_2H_2 + E_2H_2 \\
 & & E_2H_2 + E_2H_2 + E_3H_3 + E_3H_3 \\
 & & E_1H_1 + E_1H_1 + E_3H_3 + E_3H_3 \\
 & & E_2H_2 + E_2H_2 + E_1H_1 + E_1H_1 \\
 &=& 8\vec{E}\cdot\vec{H}
\end{eqnarray*}
Thus, \begin{equation}\label{c3e99}
\vec{E}\cdot\vec{H}
\end{equation}
is another invariant.

From the invariants of \eqref{c3e98} and \eqref{c3e99} we have
\begin{enumerate}
\item If the electric and magnetic fields are
perpendicular to each other in one frame of reference then they are so in all 
other inertial frames of reference.

\item If $E > H$ ($E < H$) in one frame of reference then it is so in all other
frames.

\item If the angle between $\vec{E}$ and $\vec{H}$ is acute (obtuse) in one 
frame of reference then it is so in all other frames.

\item If $\vec{E}\cdot\vec{H} \ne 0$ then we can find a frame of reference
in which the electric and magnetic fields are parallel to each other. In that
frame ${H^\op}^2 - {E^\op}^2 = H^2 - E^2$ and $\vec{E}\cdot\vec{H} = E^\op 
H^\op$.

\item if $\vec{E}\cdot\vec{H} = 0$ and if $H^2 - E^2 > 0$ then we can find
a frame in which $E^\op = 0$ and ${H^\op}^2 = H^2 - E^2 > 0$. Likewise, if $H^2
- E^2 < 0$ then we can find a frame in which $H^\op = 0$ and $-{E^\op}^2 = 
H^2 - E^2 < 0$. 
\end{enumerate}
\end{enumerate}

\section{Problems}
\begin{enumerate}
\item Find the 3-acceleration of a particle in electric and magnetic fields.
\item[Solution:] The equation of motion follows from \eqref{c3e18}, 
\eqref{c3e19} and \eqref{c3e20}. It is
\begin{equation}\label{c3e100}
\td{\vec{p}}{t} = e\vec{E} + e\frac{\vec{v}}{c}\times\vec{H}.
\end{equation}
Now, $p^i = m\gamma v^i$ and $\mathcal{E} = m\gamma c^2$ so that 
\[
p^i = \frac{\mathcal{E} v^i}{c^2}.
\]
Therefore,
\[
c^2\td{\vec{p}}{t} = \td{\mathcal{E}}{t}\vec{v} + \mathcal{E}\td{\vec{v}}{t}.
\]
From equations \eqref{c3e27} and \eqref{c3e99} we have
\[
ec^2\vec{E} + ec\vec{v}\times\vec{H} = 
e\vec{v}\cdot\vec{E}\vec{v} + \mathcal{E}\td{\vec{v}}{t}.
\]
or
\[
\td{\vec{v}}{t} = \frac{ec^2}{\mathcal{E}}
\left(\vec{E} + \frac{\vec{v}}{c}\times\vec{H} - 
\frac{\vec{v}\cdot\vec{E}}{c^2}\vec{v}\right)
= \frac{e}{m\gamma}
\left(\vec{E} + \frac{\vec{v}}{c}\times\vec{H} - 
\frac{\vec{v}\cdot\vec{E}}{c^2}\vec{v}\right)
\]
or
\begin{equation}\label{c3e101}
\td{\vec{v}}{t} = \frac{e}{m}\sqrt{1 - \frac{v^2}{c^2}}
\left(\vec{E} + \frac{\vec{v}}{c}\times\vec{H} - 
\frac{\vec{v}\cdot\vec{E}}{c^2}\vec{v}\right).
\end{equation}
In terms of $\vec{\beta}$, this equation is
\begin{equation}\label{c3e102}
\td{\vec{v}}{t} = \frac{e}{m}\sqrt{1 - \beta^2}
\left(\vec{E} + 
\vec{\beta}\times\vec{H} - (\vec{\beta}\cdot\vec{E})\vec{\beta}\right).
\end{equation}
Unlike the non-relativistic case, we cannot put acceleration as the ratio of 
force to mass. In the expression for momentum, $\gamma$ is not a constant. 
Instead of finding an expression for $d\gamma/dt$, we replace $m\gamma$ by 
$\mathcal{E}/c^2$ and use the expression for $d\mathcal{E}/dt$.

\item Maupertuis' principle for a particle in an electromagnetic field.
\item[Solution:] Maupertuis' principle states that if the energy of the 
particle is conserved then 
\[
\delta\int\vec{p}\cdot d\vec{r} = 0,
\]
where $\vec{p}$ is the generalised momentum. In the case of a charged particle 
in an electromagnetic field,
\[
\vec{p} = m\vec{v} + \frac{e}{c}\vec{A}
\]
so that the variational equation becomes
\[
\delta\int\left(m\vec{v} + \frac{e}{c}\vec{A}\right)\cdot d\vec{r} = 0 
\Rightarrow
\delta\int m\vec{v}\cdot d\vec{r} + \frac{e}{c}\delta\int\vec{A}\cdot d\vec{r} 
= 0.
\]
The energy momentum relation for a free particle is $\mathcal{E}^2 = p^2c^2 + 
m^2c^4$. For a free particle, $p = |m\vec{v}|$ and $\mathcal{E}$ is the kinetic
energy so that
\[
|m\vec{v}|^2 = \frac{\mathcal{E}^2}{c^2} - m^2c^2.
\]
For a charged particle in an electromagnetic field, the kinetic energy is, $mc^2
\gamma = \mathcal{E} - e\varphi$, according to \eqref{c3e35}. Therefore, the 
energy momentum relation is
\[
|m\vec{v}|^2 = \frac{(\mathcal{E} - e\varphi)^2}{c^2} - m^2c^2.
\]
Finally, note that $\vec{p}\cdot d\vec{r} = pdl$, where $dl$ is the length of 
the infinitesimal portion of the trajectory. This is because $d\vec{r}$ is 
parallel to the momemtum. The variational principle, thus, becomes,
\[
\delta\int 
\left(\frac{(\mathcal{E} - e\varphi)^2}{c^2} - m^2c^2\right)^{1/2}dl + 
\frac{e}{c}\delta\int\vec{A}\cdot d\vec{r} = 0.
\]

\item A charged, isotropic oscillator forced by the magnetic field.
\item[Solution:] Consider an oscillator with natural frequency $\omega_0$ and 
suppose that it obeys the equations of motion
\begin{eqnarray*}
\ddot{x} + \omega_0^2 x &=& 0 \\
\ddot{y} + \omega_0^2 y &=& 0 \\
\ddot{z} + \omega_0^2 z &=& 0
\end{eqnarray*}
Assume that the oscillator carries a charge $e$. An oscillating charge will emit
radiation and I am not sure if the these equations remain valid. But let's 
assume that the frequency of oscillation is low enough to make radiation loss 
ignorable. (I think the radiation loss can be accounted for a friction like 
term.)

Now suppose that there is a magnetic field in the $z$ axis. It exerts a force
\[
\vec{F} = \frac{e}{c}(\dot{y} H\uv{x} - \dot{x}H\uv{y})
\]
so that the equations of motion are modified to
\begin{eqnarray*}
\ddot{x} + \omega_0^2 x &=& \frac{eH}{mc}\dot{y} \\
\ddot{y} + \omega_0^2 y &=& -\frac{eH}{mc}\dot{x}\\
\ddot{z} + \omega_0^2 z &=& 0
\end{eqnarray*}
Let $\zeta = x + iy$ and write the above equations as
\begin{eqnarray*}
\ddot{x} + \omega_0^2 x &=& -\frac{ieH}{mc}\dot{(iy)} \\
\ddot{y} + \omega_0^2 y &=& -i\frac{ieH}{mc}\dot{x} \\
\ddot{(iy)} + \omega_0^2 (iy) &=& \frac{ieH}{mc}\dot{x}
\end{eqnarray*}
Adding the first and the third equations, we get
\[
\ddot{\zeta} + \frac{ieH}{mc}\zeta + \omega_0^2\zeta = 0.
\]
Let the trial solution be $\zeta = e^{\lambda t}$ so that 
\[
\lambda^2 + \frac{ieH}{mc}\lambda + \omega_0^2 = 0,
\]
the solution of which is
\[
\lambda = 
\left(-\frac{ieH}{mc} \pm \sqrt{-\frac{e^2H^2}{m^2c^2} - 4\omega_0^2}\right)
\frac{1}{2} = -i\left(\frac{eH}{2mc} \pm 
\sqrt{\omega_0^2 + \frac{e^2H^2}{4m^2c^2}}\right).
\]
Therefore, the frequency of the oscillator is
\[
-i\lambda = \omega = \sqrt{\omega_0^2 + \frac{e^2H^2}{4m^2c^2}} \pm 
\frac{eH}{2mc}.
\]
If the field is weak, that is if
\[
H \ll \frac{2mc}{e},
\]
we can ignore the second term under the square root and
\[
\omega \approx \omega_0 \pm \frac{eH}{2mc}.
\]

\item Relativistic motion of a charged particle in parallel, uniform electric
and magnetic fields.
\item[Solution:] Let $\vec{E} = E\uv{z}, \vec{H} = H\uv{z}$ so that the 
equations of motion are
\[
\td{\vec{p}}{t} = eE\uv{z} + \frac{e}{c}\vec{v}\times\vec{H}
\]
that is,
\begin{eqnarray}
\td{p_x}{t} &=& \frac{e}{c}v_yH \label{c3e103} \\
\td{p_y}{t} &=& -\frac{e}{x}v_xH \label{c3e104} \\
\td{p_z}{t} &=& eE \label{c3e105} 
\end{eqnarray}
Equation \eqref{c3e105} can be readily integrated to
\[
p_z(t) = eEt + p_z(0).
\]
Since $\vec{p} = m\gamma\vec{v}$, we have
\begin{equation}\label{c3e106} 
v_z = \frac{eE}{m\gamma}t + \frac{p_z(0)}{m\gamma}.
\end{equation}
We choose $t = 0$ such that $p_z(0) = 0$ and use the relation, 
$\mathcal{E} = m\gamma c^2$ so that
\begin{equation}\label{c3e107}
v_z = \frac{eEc^2}{\mathcal{E}}t.
\end{equation} 
From equation \eqref{c3e27},
\[
\td{\mathcal{E}}{t} = ev_zE
\]
so that, using \eqref{c3e106}, we have 
\[
\td{\mathcal{E}}{t} = (eEc)^2\frac{t}{\mathcal{E}} \Rightarrow
\frac{\mathcal{E}^2}{2} = \frac{(eEct)^2}{2} + \frac{\mathcal{E}_0^2}{2},
\]
so that, we have
\begin{equation}\label{c3e108}
\mathcal{E} = \sqrt{\mathcal{E}_0^2 + (eEct)^2}.
\end{equation}
Substituting this in \eqref{c3e106},
\[
\dot{z} = c\frac{eEct}{\sqrt{\mathcal{E}_0^2 + (eEct)^2}}
\]
To solve this equation, put $eEct = \mathcal{E}_0\sinh u$ so that $eEcdt = 
\mathcal{E}_0\cosh udu$ and
\[
dz = \frac{\mathcal{E}_0}{eE}\sinh udu \Rightarrow
z(t) = z(0) + \frac{\mathcal{E}_0}{eE}\sinh u = z(0) + \frac{\mathcal{E}_0}{eE}
\sqrt{1 + \frac{(eEc)^2}{\mathcal{E}_0^2}t^2}
\]
If $z(0) = 0$ then 
\begin{equation}\label{c3e109}
z(t) = \frac{\mathcal{E}_0}{eE}\sqrt{1 + \frac{(eEc)^2}{\mathcal{E}_0^2}t^2}.
\end{equation}
We now turn our attention to the $xy$ plane. From equations \eqref{c3e102} and
\eqref{c3e103},
\[
\frac{d}{dt}(p_x + ip_y) = \frac{eH}{c}(v_y - iv_x) = 
-\frac{ieH}{c}(v_x + iv_y).
\]
Since $\vec{p} = (\mathcal{E}/c^2)\vec{v}$, the above equation becomes
\begin{equation}\label{c3e110}
\frac{d}{dt}(p_x + ip_y) = -\frac{ieHc}{\mathcal{E}}(p_x + ip_y).
\end{equation}
Now introduce a variable $\phi$ such that
\begin{equation}\label{c3e111}
d\phi = \frac{eHc}{\mathcal{E}}dt,
\end{equation}
so that the solution of \eqref{c3e109} can be written as
\begin{equation}\label{c3e112}
p_x + ip_y = p_t e^{-i\phi},
\end{equation}
where $p_t$ is a constant of integration. Putting \eqref{c3e107} in 
\eqref{c3e110}, we get
\begin{equation}\label{c3e113}
\phi = eHc\frac{dt}{\sqrt{\mathcal{E}_0^2 + (eEct)^2}} = \frac{H}{E}\sinh^{-1}
\left(\frac{eEct}{\mathcal{E}_0}\right),
\end{equation}
which can be written as
\begin{equation}\label{c3e114}
ct = \frac{\mathcal{E}_0}{eE}\sinh\left(\frac{E}{H}\phi\right).
\end{equation}
Since $\vec{p} = (\mathcal{E}/c^2)\vec{v}$, we can write \eqref{c3e111} as
\[
(\mathcal{E}/c^2)(v_x + iv_y) = p_t e^{-i\phi} \Rightarrow 
(\mathcal{E}/c^2)\frac{d}{dt}(x + iy) = p_te^{-i\phi}
\]
From \eqref{c3e110}, we get
\[
d(x + iy) = \frac{cp_t}{eH}e^{-i\phi}d\phi \Rightarrow x + iy = 
\frac{icp_t}{eH}e^{-i\phi}
\]
or
\[
x + iy = \frac{cp_t}{eH}\sin\phi + i\frac{cp_t}{eH}\cos\phi
\]
from which we get
\begin{eqnarray}
x(\phi) &=& \frac{cp_t}{eH}\sin\phi \label{c3e115} \\
y(\phi) &=& \frac{cp_t}{eH}\cos\phi \label{c3e116}
\end{eqnarray}
The equation of the trajectory is
\begin{eqnarray}
x(t) &=& 
\frac{cp_t}{eH}\sin\left(
\frac{H}{E}\sinh^{-1}\left(\frac{eEct}{\mathcal{E}_0}\right)\right) 
\label{c3e117} \\
y(t) &=& \frac{cp_t}{eH}\cos\left(
\frac{H}{E}\sinh^{-1}\left(\frac{eEct}{\mathcal{E}_0}\right)\right) 
\label{c3e118} \\
z(t) &=& \frac{\mathcal{E}_0}{eE}\sqrt{1 + \frac{(eEc)^2}
{\mathcal{E}_0^2}t^2}. \label{c3e119}
\end{eqnarray}
\end{enumerate}

