\documentclass{beamer}
\title{Electricity and Magnetism}
\subtitle{Introduction to the course}
\author{Amey Joshi}

\begin{document}
\frame{\titlepage}

\begin{frame}
\frametitle{Why this course?}
\begin{enumerate}
\item Some of us did not get a chance to study the subject at an under-grad level.
\item As a result, we find it hard to formulate problems although we understand the
ideas.
\item We will start at a more elementary level, build our intuition, solve a lot
of problems and prepare ourselves to tackle the graduate level material. 
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Why Purcell's book?}
\begin{enumerate}
\item It starts at an under-grad level, is not too demanding on mathematical techniques
and yet addresses many advanced topics.
\item It is full of insights. It tells many things that other books miss, some quite
deep.
\item Like Feynman's and Landau \& Lifshitz's books, this one shows how great physicist
solves problem. Often times, they arrive at a rough solution through simple physics-based
arguments. They use maths to refine it.
\item Edward Purcell was a Noble laureate who took time to write a book for the
undergrads.
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Our plan}
\begin{enumerate}
\item We will cover the material in roughly 24 weeks.
\item We will spend rought 2/3rd of our time together solving problems.
\item We will present a few key points of the chapter and solve problems related
to them.
\item We are using the second edition of the book. It will help us to be bi-lingual
in gaussian and SI units.
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{E\&M with friends}
\begin{enumerate}
\item This is a group of friends. Don't hesitate to ask questions.
\item Sometimes we will not be able to solve a problem or answer a question. That
just makes the subject more interesting.
\item Please read the book carefully. It tells far more than other books at its
level.
\item Participate in the discussion when we solve problems. We will solve a few
every once in a while.
\end{enumerate}
\end{frame}

\end{document}